---
title: "Reproduceable research Course project 1"
author: "Jacob Mokgadi"
date: "25/01/2022"
output: html_document
---



## Load the data
Load the data

```{r}
get.data <- function (){
        if (!file.exists("data")) {
        message("Creating Data folder in working directory")
        dir.create("data")
        }

        if(!file.exists("data/repdata-data-activity")) {
        fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileURL, destfile = "./data/repdata-data-activity.zip")
        file.name <- "./data/repdata-data-activity.zip"
        unzip(file.name, exdir = "data")
        unlink(file.name)
}
else message("data already exists")
}
get.data()
activity <- read.csv("./data/activity.csv", colClasses = c("numeric", "Date", "numeric"))
activity$day <- weekdays(activity$date)
summary(activity)
```
## Steps taken per day

```{r}
stepsPerDay <- aggregate(steps ~ date, activity, sum, na.rm=TRUE)
stepsPerDay
```
## Histogram of the total number of steps taken per day

```{r}
require (ggplot2)
activityAgg <- aggregate(steps ~ date, activity, sum, na.rm = TRUE)
qplot(steps, data = activityAgg, binwidth = 1400) +
theme_bw()
```

Mean Steps per day
```{r}
mean(activityAgg$steps)
```

Median steps per day
```{r}
median(activityAgg$steps)
```

## Average Daily activity
Time series plot of the 5-minute interval and the average number of steps taken.

```{r}
intervalAve <- aggregate(steps ~ interval, activity, mean, na.rm = TRUE)
ggplot(intervalAve, aes(x=interval,y=steps)) + 
geom_line(color="purple",size=1) +  
labs(x="Interval",y="Average Number of Steps") +
theme_bw()
```

## 5-minute interval containing the maximum number of steps
```{r}
intervalAve[which.max(intervalAve$steps),]$interval
```

## Inputing Missing Values

Number of missing values
```{r}
sum(is.na(activity))
```

## Method for inputing missing steps
Missing step values (NA) were replaced by the mean number of steps taken for the corresponding time interval and weekday. 

## create new dataset using the imputed values
```{r}
intdayAve <- aggregate(steps ~ interval + day, activity, mean, na.rm = TRUE)
activityImpute <- merge(activity, intdayAve, by=c("interval", "day"))
activityImpute <- transform(activityImpute, steps.x = ifelse(is.na(steps.x),steps.y,steps.x))
activityImpute <- data.frame(activityImpute[,1:4])
names(activityImpute) <- c("interval", "day","steps", "date")
activityImpute$steps <- round(activityImpute$steps, digits = 0)
activityImpute <- activityImpute[order(activityImpute$date, activityImpute$interval),]
```

## Histogram of total number of steps taken per day, using new dataset

```{r}
activityImputeAgg <- aggregate(steps ~ date, activityImpute, sum, na.rm = TRUE)
qplot(steps, data = activityImputeAgg, binwidth = 1400) +
theme_bw()
```

## Mean steps per day with inputed values
```{r}
mean(activityImputeAgg$steps)
```

## Median steps per day with inputed values
```{r}
median(activityImputeAgg$steps)
```
Including the inputed values in the dataset, both the median and the mean total number of steps taken per day increase, as expected


## Investigating Differences in Activity Patterns
```{r}
activityImpute$daytype <- ifelse(activityImpute$day %in% c("Saturday", "Sunday"),"Weekend", "Weekday")
activityImputeDayAgg <- aggregate(steps ~ interval + daytype, activityImpute, mean)
ggplot(activityImputeDayAgg, aes(x=interval,y=steps)) + 
geom_line(color="purple",size=1) + 
facet_wrap(~daytype, nrow=2, ncol=1) + 
labs(x="Interval",y="Number of Steps") +
theme_bw()
```






